MATH 245 Homework 5

### 2024-04-07 PDEs/math245_hw5.Rmd
pde hw 5 prob 5 and 6

---
title: "MATH 245 Homework 5"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-04-05"
output:
  pdf_document:
    extra_dependencies: ["tcolorbox", "xcolor"]
---

@@

## Question 5: Radioactive Decay Problem

\begin{equation}\label{eqn:radio} f(x)=
    \begin{cases}
     u_t - u_{xx} = Ae^{-ax} \\
     u(0,x) = \sin{x} \\
     u(t, 0) = u(t, \pi) = 0
    \end{cases}       
\end{equation}

@@

From $b'_0(t)=q_0(t)$ we get $b_0(t)=\int_{0}^{t} q_0(s)\;ds$ Since $q_0=\frac{-A}{a \pi} \left(e^{-a\pi}-1\right)$, this means $b_0(t)=\frac{-At}{a \pi} \left(e^{-a\pi}-1\right)$

@@

Therefore, our solution to (\ref{eqn:radio}) is:
$$u(t,x)= b_0(t)+\sum_{n=1}^{\infty} b_n(t) \sin{(nx)}, \quad \text{where} \\ b_n(t)=e^{-n^2}b_n(0)+\int_{0}^{t} e^{n^2(s-t)}q_n\;ds,\\
q_n=\frac{2A}{\pi} \frac{n(-1)^{n+1} e^{-a\pi}+n}{n^2+a^2}, \\ 
b_0(t)=\frac{-At}{a \pi} \left(e^{-a\pi}-1\right)$$

$$u(0,x)=b_0(0)+\sum_{n=1}^{\infty} \left[e^{-n^2}b_n(0)+\int_{0}^{0} e^{n^2(s-0)}q_n(s)\;ds   \right] \sin{(nx)}$$

$$u(0,x)=b_0(0)+\sum_{n=1}^{\infty}e^{-n^2}b_n(0) \sin{(nx)}$$

>>>>

---
title: "MATH 245 Homework 5"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-04-07"
output:
  pdf_document:
    extra_dependencies: ["tcolorbox", "xcolor"]
---

@@

## Problem 1: Inhomogeneous Heat Equation

Using the method of separation of variables, solve the inhomogeneous heat equation:
\begin{equation}\label{eqn:inhomoheat}
    \begin{cases}
     u_{t} -k u_{xx}=x \cos{t}          & 0<x<\pi, \quad t>0 \\
     u(0,x) = \sin{(\pi x)}             & 0<x<\pi\\
     u(t,0)=t^2, \;\; u(t,\pi)=2t       & t>0
    \end{cases}       
\end{equation}

## Problem 2: Inhomogeneous Wave Equation

Using the method of separation of variables, solve the inhomogeneous wave equation:
\begin{equation}\label{eqn:inhomowave}
    \begin{cases}
     u_{tt} -c^2 u_{xx}=F(t,x)    & 0<x<L, \quad t>0 \\
     u(0,x) = \phi(x)             & 0<x<L\\
     u_t(0,x) = \psi(x)           & 0<x<L\\
     u_x(t,0)=h(t), \;\; u_x(t,L)=g(t) & t>0
    \end{cases}       
\end{equation}

for constants $a$ and $k>0$

## Problem 3: Damped Heat Equation

Using the method of separation of variables, solve the damped heat equation:
\begin{equation}\label{eqn:dampheat}
    \begin{cases}
     u_{t} +au=ku_{xx}    & -\pi<x<\pi, \quad t>0 \\
     u(0,x) = \phi(x)     & -\pi<x<\pi\\
     u(t,\pi)=u(t,-\pi)   & t>0 \\
     u_x(t,\pi)=u_x(t,-\pi) & t>0
    \end{cases}       
\end{equation}

for constants $a$ and $k>0$

## Problem 4: Beam Equation

Using the method of separation of variables, solve the beam equation:
\begin{equation}\label{eqn:beam}
    \begin{cases}
     u_{tt} =c^2 u_{xxxx} & 0<x<L, \quad t>0 \\
     u(0,x) = \phi(x)     & 0<x<L\\
     u_t(0,x) = \psi(x)   & 0<x<L\\
     u(t, 0) = u(t, L) = 0 & t>0 \\
     u_{xx}(t,0)=u_{xx}(t,L)=0
    \end{cases}       
\end{equation}

## Problem 5: Radioactive Decay Problem

Using the method of separation of variables, solve the radioactive decay problem, for constants $A, a>0$.
\begin{equation}\label{eqn:radio}
    \begin{cases}
     u_t - u_{xx} = Ae^{-ax} \\
     u(0,x) = \sin{x} \\
     u(t, 0) = u(t, \pi) = 0
    \end{cases}       
\end{equation}

@@

From $b'_0(t)=q_0(t)$ we get $$b_0(t)=\int_{0}^{t} q_0(s)\;ds$$

Since $q_0=\frac{-A}{a \pi} \left(e^{-a\pi}-1\right)$, this means
$$b_0(t)=\frac{-At}{a \pi} \left(e^{-a\pi}-1\right)+b_0(0)$$

@@

$$u(0,x)= \sin{x} = b_0(0)+\sum_{n=1}^{\infty} b_n(0) \sin{(nx)}$$

so using our equations to find the  coefficients of a Fourier sine series, 
$$b_0(0)=\frac{2}{\pi} \int_{0}^{\pi}\sin{(nx)}dx$$
$$=\frac{-2}{n\pi} \cos{(nx)}\Bigr\rvert^{\pi}_{0}\;dx\\
=  \frac{2}{n\pi}(1+(-1)^{n+1})$$

$$b_n(0)=\frac{1}{\pi} \int_{0}^{\pi}\sin{(x)} \sin{(nx)}dx$$

We evaluate this integral using the following trigonometric identity: $$\sin{\alpha}\sin{\beta}=\frac{\cos{(\alpha-\beta)}-\cos{(\alpha+\beta)}}{2}
\quad \Longrightarrow\quad \sin{(x)} \sin{(nx)}=\frac{\cos{(-x)}-\cos{(3x)}}{2}$$

$$b_n(0)=\frac{1}{2\pi} \int_{0}^{\pi} \cos{(x)}-\cos{(3x)}dx=0$$
Because $\sin{0}, \sin{\pi}$, and $\sin{3\pi}$ all equal zero.

Therefore, our solution to (\ref{eqn:radio}) is:
\begin{tcolorbox}[colback=white, title=Solution]
\begin{align*}
& u(t,x)= b_0(t)+\sum_{n=1}^{\infty} b_n(t) \sin{(nx)}, \quad \text{where} \\ 
& b_n(t)=\int_{0}^{t} e^{n^2(s-t)}q_n\;ds,\\
& q_n=\frac{2A}{\pi} \frac{n(-1)^{n+1} e^{-a\pi}+n}{n^2+a^2}, \\ 
& b_0(t)=\frac{-At}{a \pi} \left(e^{-a\pi}-1\right)+b_0(0), \\
& b_0(0)=\frac{2}{n\pi}(1+(-1)^{n+1})
\end{align*}
\end{tcolorbox}

## Problem 6: Telegraph Equation

Using the method of separation of variables, solve the telegraph equation:
\begin{equation}\label{eqn:tele}
    \begin{cases}
     u_{tt} + au_t + bu = c^2 u_{xx}   & 0<x<l, \quad t>0 \\
     u(0,x) = \phi(x)             & 0 \leq x \leq l\\
     u_t(0,x) = \psi(x)           & 0 \leq x \leq l\\
     u(t,0)=u(t,l)=0 & t>0
    \end{cases}       
\end{equation}

for constants $a, b >0$. Only find the solution when the characteristic equation of the time problem has real roots. Define the following energy:

$$E(t)=\frac{1}{2} \int_{0}^{l}(u_t^2+c^2u_x^2+bu^2)\;dx $$
Show that $E(t) \leq E(0)$ for all $t>0$. Then prove that the telegraph equation has a unique solution.

- - - -

Suppose we have a separated solution of the form $u(t,x)=X(x)T(t)$.

If we plug this into the homogeneous Dirichlet BCs, we find $u(t,0)=T(t)X(0)=0, u(t,l)=T(t)X(l)=0$. In order for both of these to be true, we must have either $T(t)=0$ or $X(0)=X(l)=0$.

But if $T(t)=0$, then $u(t,x)=0$ for all $x$, which contradicts our ICs. Thus, we must have $X(0)=X(l)=0$.

Plugging our expression $u(t,x)=X(x)T(t)$ into the PDE in (\ref{eqn:tele}), it becomes $$T''(t)X(x)+aT'(t)X(x)+bT(t)X(x)=c^2T(t)X''(x)$$
$$- \frac{T''(t)+aT'(t)+bT(t)}{c^2 T(t)}=-\frac{X''(x)}{X(x)}=\lambda$$

Based on our assumptions thus far and the IBVP (\ref{eqn:tele}), we have three problems:

**Spatial problem**
\begin{align*} \begin{cases}
X''+\lambda X=0 \\
X(0)=X(l)=0 \end{cases} \end{align*}

**Time problem**
\begin{equation}\label{eqn:time6}T''+aT'+(b+\lambda c^2) T=0\end{equation}

**IVP**
$$u(0,x) = T(0)X(x)=\phi(x)$$

The spatial problem is an eigenvalue problem with homogenous Dirichlet BCs, which we already know has the eigenvalues and eigenvectors
$$\lambda_n= \left(\frac{n\pi}{l}\right)^2, \quad X_n(x)=\sin{\left( \frac{n\pi x}{l}\right)},\quad n=1,2,3... $$

Now moving to the time problem, the equation (\ref{eqn:time6}) becomes the characteristic equation $r^2 + ar + (b+\lambda c^2)=0$. From the Pythagorean Theorem, the roots of this equation are $$\frac{-a\pm\sqrt{a^2-4(b+\lambda c^2)}}{2}$$

In order for the characteristic equation of the time problem to have only real roots, we must have $a^2-4(b+\lambda c^2) \geq 0$, or $$\lambda \leq \frac{a^2-4b}{4c^2}$$

For each $\lambda_n$ associated with the spatial problem, we get a solution to the time problem:
$$T_n(t)=C_n \exp{\left[\frac{t}{2}(-a-a^2-4b-4\lambda_n c^2)\right]}+D_n \exp{\left[\frac{t}{2}(-a+a^2-4b-4\lambda_n c^2)\right]} $$

where $C_n, D_n \in \mathbb{R}$

Therefore, the following are solutions of the PDE in (\ref{eqn:tele}):
$u_n(t,x)=X_n(x)\;T_n(t), \quad n=1,2,3...$

By the principle of superposition,
\begin{align*}
u(t,x) &= \sum^{\infty}_{n=1} c'_n \;u_n(t,x) \\
       &=\sum^{\infty}_{n=1} c'_n \;T_n(t)\;X_n(x) \\
       &=\sum^{\infty}_{n=1} A_n \;\exp{\left[\frac{t}{2}(-a-a^2-4b-4\lambda_n c^2)\right]}\;\sin{\left( \frac{n\pi x}{l}\right)}+ \\
       &\sum^{\infty}_{n=1} B_n \;\exp{\left[\frac{t}{2}(-a+a^2-4b-4\lambda_n c^2)\right]}\;\sin{\left( \frac{n\pi x}{l}\right)}
\end{align*}

where $A_n = c'_n C_n$ and $B_n = c'_n D_n$, with $A_n, B_n \in \mathbb{R}$.

We can get rid of these arbitrary constants by using our initial conditions,  and . First, let us simplify the notation by defining $\gamma=\frac{t}{2}(-a-a^2-4b-4\lambda_n c^2)$ and $\zeta=\frac{t}{2}(-a-a^2-4b-4\lambda_n c^2)$.

Using our first initial condition, $u(0,x) = \phi(x)$:
$$u(t,x)= \sum^{\infty}_{n=1} A_n \,e^{\gamma t} \sin{\left( \frac{n\pi x}{l}\right)}+ \\
\sum^{\infty}_{n=1} B_n \,e^{\zeta t} \sin{\left( \frac{n\pi x}{l}\right)}$$

$$u(0,x) = \phi(x) = \sum^{\infty}_{n=1} \left(A_n+B_n\right) \sin{\left( \frac{n\pi x}{l}\right)}$$

We can use the equation for the coefficients inside a Fourier sine series to find $A_n+B_n$
$$ A_n+B_n=\frac{2}{l} \int_{0}^{l}\phi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx$$

Now repeating for our second initial condition, $u_t(0,x) = \psi(x)$:
$$u_t(t,x)= \sum^{\infty}_{n=1} \gamma\, A_n \,e^{\gamma t} \;\sin{\left( \frac{n\pi x}{l}\right)}+  \sum^{\infty}_{n=1} \zeta\,B_n \,e^{\zeta t} \;\sin{\left( \frac{n\pi x}{l}\right)}$$

$$u_t(0,x)= \psi(x) = \sum^{\infty}_{n=1} \left(\gamma A_n+\zeta B_n\right) \sin{\left( \frac{n\pi x}{l}\right)}$$

$$\left(\gamma A_n+\zeta B_n\right)=\frac{2}{l} \int_{0}^{l}\psi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx$$
Which can be rearranged as
$$A_n= \frac{1}{\gamma}\left[-\zeta B_n +\frac{2}{l} \int_{0}^{l}\psi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx \right]$$

Now we can solve for $B_n$:

$$\frac{-\zeta B_n}{\gamma}+\frac{2}{\gamma l} \int_{0}^{l}\psi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx= -B_n+\frac{2}{l} \int_{0}^{l}\phi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx$$

$$B_n-\frac{\zeta B_n}{\gamma}= \left(1-\frac{\zeta}{\gamma}\right)B_n=
\frac{2}{l} \int_{0}^{l}\phi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx-\frac{2}{\gamma l} \int_{0}^{l}\psi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx$$

Therefore, our solution is:
\begin{tcolorbox}[colback=white, title=Solution]
$$u(t,x)= \sum^{\infty}_{n=1} A_n \,e^{\gamma t} \sin{\left( \frac{n\pi x}{l}\right)}+ \\
\sum^{\infty}_{n=1} B_n \,e^{\zeta t} \sin{\left( \frac{n\pi x}{l}\right)}, \quad \text{where}$$

$$ A_n=\frac{2}{l} \int_{0}^{l}\phi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx-B_n$$
$$B_n=\frac{2\gamma}{(\gamma-\zeta)l} \int_{0}^{l}\phi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx-\frac{2(\gamma-\zeta)}{l} \int_{0}^{l}\psi(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx,$$
$$\gamma=\frac{t}{2}(-a-a^2-4b-4\lambda_n c^2)$$
$$\zeta=\frac{t}{2}(-a-a^2-4b-4\lambda_n c^2)$$
$$\lambda_n= \left(\frac{n\pi}{l}\right)^2, \quad n=1,2,3...$$
\end{tcolorbox}
